---
title: "Predicting High-Potential Restaurant Locations in Bangladesh"
subtitle: "Introduction to Data Science — Semester Project (Phase 1: Midterm)"
author: "Md.Daud Ibrahim (ID: 22-49760-3)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(42)
```

# 1. Install and Load Packages

We will check, install, and load all the required packages for the project using the standard loop structure from the labs:

```{r packages, message=FALSE, warning=FALSE}
required_packages <- c(
  "tidyverse",
  "skimr",
  "moments",
  "corrplot",
  "knitr",
  "kableExtra",
  "e1071",
  "caret"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE)) {
    cat("Installing", pkg, "...\n")
    install.packages(pkg, repos = "https://cran.r-project.org")
    library(pkg, character.only = TRUE)
  } else {
    cat(pkg, "is already installed.\n")
  }
}
```

---

# Section 1 — Problem Framing

## 1.1 Domain Context and Dataset Background
The restaurant business in Bangladesh is growing very fast, especially in big cities like Dhaka and Chittagong. Many new restaurants are opening, but a lot of them also close down quickly. One of the main reasons for this failure is choosing the wrong location. Setting up a restaurant needs a lot of investment (for decoration, rent, and staff), so picking the right place is very important for the business to survive and make profit.

In this project, we are using the **BanglaDine2024 Bangladeshi Restaurant Industry Dataset** to analyze restaurant locations. This dataset has information on 58,376 restaurants collected from Google Maps, including coordinates (lat, lng), division, restaurant types, customer ratings, review counts, and service options (like dine-in, delivery, takeaway). Our goal is to build a machine learning model to predict which locations have high potential for opening a new restaurant.

## 1.2 Analytical Questions
To guide our analysis and project workflow, we want to answer these three main questions:

1. **Do specific divisions or coordinates in Bangladesh have a higher chance of having successful restaurants?**
   *We want to see if opening in certain regions (like Dhaka or Chittagong) automatically gives a higher chance of success.*
2. **How does having different service options like delivery and takeaway affect restaurant popularity?**
   *We will check if offering multiple options (delivery, takeaway, dine-in) makes a restaurant more popular.*
3. **Does having more competitors nearby help a restaurant or make it fail?**
   *We will analyze if a high number of restaurants in the same area is good (food hubs that attract customers) or bad (too much competition).*

## 1.3 Success Criteria and Evaluation Metrics
*(Skip it for mid-term per faculty instructions)*

## 1.4 Dataset Justification
This dataset is very suitable for our project because:
* It has exact latitude and longitude coordinates, which lets us do mapping and spatial analysis.
* It is a large dataset with over 51,000 clean records covering all 8 administrative divisions of Bangladesh, so it represents the whole country.
* It contains real customer ratings and review counts from Google Maps, which are good measures of restaurant success.

---

# Section 2 — Data Acquisition & Initial Inspection

## 2.1 Loading the Data
We load the dataset using the raw GitHub URL. We also included a check for the local file so it is easy to run and compile on any machine.

```{r load-data}
# Explicitly load tidyverse to make the chunk self-contained
library(tidyverse)

# Public URL of the dataset on GitHub
url <- "https://raw.githubusercontent.com/MdDaudIbrahim/Data-Science-Summer-2025-2026/main/DATASET/BanglaDine2024%20Bangladeshi%20Restaurant%20Industry%20Dataset/Restaurant_Spatial-Analytical.csv"

# Local path
local_path <- "DATASET/BanglaDine2024 Bangladeshi Restaurant Industry Dataset/Restaurant_Spatial-Analytical.csv"

if (file.exists(local_path)) {
  df <- readr::read_csv(local_path, show_col_types = FALSE)
} else {
  df <- readr::read_csv(url, show_col_types = FALSE)
}

# Filter for Bangladesh divisions only to clean foreign records
bd_divisions <- c("Dhaka", "Chittagong", "Khulna", "Rajshahi", "Rangpur", "Sylhet", "Barisal", "Mymensingh")
df <- df %>% 
  filter(division %in% bd_divisions)

# Clean all character columns to contain only valid ASCII characters to prevent locale/gsub errors
df <- df %>%
  mutate(across(where(is.character), ~ iconv(., to = "ASCII", sub = "")))

# Preview the first few rows
head(df, 5)
```

**Observation:** The preview shows standard columns from Google Maps like name, lat, lng, rating, and reviews. Some rows have NA values in rating and reviews, meaning they are unrated.

## 2.2 Dimensions and Variable Types
We inspect the dataset size and variables to understand what data types we are dealing with.

```{r inspect}
dim(df)
str(df %>% select(name, lat, lng, division, types, rating, user_ratings_total, `Service options`))
```

**Observation:** The filtered dataset has **51,277 rows** and **211 columns**. The coordinates are numeric (`dbl`), and columns like name, division, and types are character text (`chr`).

## 2.3 Summary Statistics
We run summary statistics to check the range and mean of the columns.

```{r summary}
summary(df %>% select(lat, lng, rating, user_ratings_total, types, division))
```

**Observation:** 
* The median rating is 4.1, which means most restaurants have good ratings.
* The review count (`user_ratings_total`) goes from 1 to 27,484, but the median is only 5. This shows the data is highly skewed.
* There are 11,868 missing values in ratings and reviews.

## 2.4 Data Provenance
* **Source:** Google Maps API extraction (BanglaDine2024 dataset).
* **Collection method:** Scraped listings from food-related business searches across Bangladesh divisions in late 2024.
* **Known limitations:** New or very small restaurants do not have ratings. The data also has more listings in big cities compared to rural areas.

---

# Section 3 — Exploratory Data Analysis (EDA)

We define a binary target variable **`is_high_potential`** indicating restaurant success: rating $\ge 4.0$ and review volume $\ge$ the median of all active reviewed restaurants.

```{r define-target}
median_reviews <- median(df$user_ratings_total, na.rm = TRUE)

df <- df %>%
  mutate(is_high_potential = ifelse(!is.na(rating) & rating >= 4.0 & user_ratings_total >= median_reviews, 1, 0))
```

## 3.1 Univariate Analysis — Numeric Variables
We use histograms to check the distribution of ratings and reviews.

**Rating Distribution (Base R style):**
```{r univariate-numeric-base}
hist(
  df$rating,
  main = "Distribution of Restaurant Ratings",
  xlab = "Google Maps Rating",
  col = "lightblue",
  border = "black"
)
```

**Reviews Distribution (ggplot2 style):**
```{r univariate-numeric-gg}
ggplot(
  df,
  aes(x = user_ratings_total)
) +
  geom_histogram(
    bins = 30,
    fill = "steelblue",
    color = "black"
  ) +
  scale_x_log10(labels = scales::comma) +
  labs(
    title = "Distribution of Review Counts (Log Scale)",
    x = "Reviews (Log10 scale)",
    y = "Frequency"
  )
```

**Observation:** The ratings are left-skewed, showing most rated restaurants get 3.5 to 4.8 stars. The review counts are highly right-skewed; most places have very few reviews, while a few popular ones have thousands.

## 3.2 Univariate Analysis — Categorical Variables
We check the number of restaurants in each division using a bar chart.

```{r univariate-categorical}
df %>%
  count(division) %>%
  ggplot(
    aes(
      x = reorder(division, n),
      y = n,
      fill = division
    )
  ) +
  geom_col(alpha = 0.85) +
  coord_flip() +
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "Restaurant Counts by Division",
    x = "Division",
    y = "Number of Restaurants"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Observation:** Dhaka has the highest number of restaurants (around 20,000), followed by Chittagong (11.5k). Barisal and Mymensingh have the lowest counts.

## 3.3 Skewness Analysis
We calculate the skewness of the numeric variables to see if they are symmetric or skewed.

```{r skewness}
skew_vals <- df %>%
  select(rating, user_ratings_total, lat, lng) %>%
  summarise(across(everything(), ~round(e1071::skewness(., na.rm = TRUE), 3))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Skewness") %>%
  mutate(
    Direction = case_when(
      Skewness >  1   ~ "Highly right-skewed",
      Skewness >  0.5 ~ "Moderately right-skewed",
      Skewness < -1   ~ "Highly left-skewed",
      Skewness < -0.5 ~ "Moderately left-skewed",
      TRUE            ~ "Approximately symmetric"
    )
  ) %>%
  arrange(desc(abs(Skewness)))

knitr::kable(skew_vals, caption = "Skewness of Numeric Variables") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

**Observation:** The review count column is highly right-skewed (skewness is 16.6), which means it needs log transformation. Rating is moderately left-skewed, and coordinates are mostly symmetric.

## 3.4 Bivariate Analysis — Predictors vs. Target
We check the percentage of high-potential restaurants in each division.

```{r bivariate}
df %>%
  group_by(division) %>%
  summarise(high_potential_pct = mean(is_high_potential) * 100) %>%
  ggplot(
    aes(
      x = reorder(division, high_potential_pct),
      y = high_potential_pct,
      fill = division
    )
  ) +
  geom_col(alpha = 0.85) +
  coord_flip() +
  scale_fill_brewer(palette = "Oranges") +
  labs(
    title = "High-Potential Restaurant Percentage by Division",
    x = "Division",
    y = "Percentage (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Observation:** Sylhet division has the highest percentage of successful restaurants (over 60%), followed by Chittagong and Dhaka. This could be because Sylhet and Chittagong (Cox's Bazar) are tourist spots with highly-rated options.

## 3.5 Correlation Analysis
We calculate the correlation between variables and plot it using the `corrplot` package.

```{r correlation}
cor_data <- df %>%
  select(rating, user_ratings_total, lat, lng) %>%
  drop_na()

cor_matrix <- cor(cor_data)

corrplot::corrplot(
  cor_matrix,
  method = "color",
  type = "upper"
)
```

**Observation:** The correlation between ratings and reviews is very weak (0.1). Coordinates do not show any linear correlation with rating.

## 3.6 Notable Finding: Spatial Clustering of High-Potential Locations
We plot the coordinates of restaurants in Dhaka and Chittagong to see their distribution.

```{r notable-finding}
df %>%
  filter(division %in% c("Dhaka", "Chittagong")) %>%
  ggplot(
    aes(
      x = lng,
      y = lat,
      color = factor(is_high_potential)
    )
  ) +
  geom_point(alpha = 0.3, size = 1) +
  scale_color_manual(
    values = c("orange", "steelblue"),
    labels = c("Low Potential", "High Potential")
  ) +
  labs(
    title = "Spatial Clustering in Dhaka & Chittagong",
    x = "Longitude",
    y = "Latitude",
    color = "Potential Class"
  ) +
  theme_minimal()
```

**Observation:** The map shows that high-potential restaurants (blue points) are clustered close together in city centers. This tells us that opening a restaurant near existing successful spots is generally better than in isolated areas.

---

# Section 4 — Preprocessing & Feature Engineering

## 4.1 Missing Values
We check the counts and percentages of missing values.

```{r missing-values}
colSums(is.na(df %>% select(name, lat, lng, division, types, rating, user_ratings_total)))

round(
  colSums(is.na(df %>% select(name, lat, lng, division, types, rating, user_ratings_total))) /
  nrow(df) * 100,
  2
)
```

We remove rows with missing ratings or reviews. Missing names are set to "Unknown".

```{r impute}
df <- df %>% 
  filter(!is.na(rating) & !is.na(user_ratings_total))

df$name[is.na(df$name)] <- "Unknown"
```

**Decision:** We deleted rows with missing ratings and reviews because if they don't have reviews, we cannot label them as successful or unsuccessful. Imputing these target values would make our data inaccurate.

## 4.2 Outlier Detection and Treatment
We check the outliers in review counts using the IQR method.

```{r outliers}
Q1 <- quantile(df$user_ratings_total, 0.25)
Q3 <- quantile(df$user_ratings_total, 0.75)
IQR_value <- IQR(df$user_ratings_total)

lower <- Q1 - 1.5 * IQR_value
upper <- Q3 + 1.5 * IQR_value

outliers <- df$user_ratings_total[
  df$user_ratings_total < lower |
  df$user_ratings_total > upper
]

cat("Total Outliers identified by IQR:", length(outliers), "\n")
```

**Decision:** Removing outliers using the IQR method would delete highly popular and famous restaurants (which are our best examples of success). Instead, we apply a log transformation `log(x + 1)` to handle the extreme values.

```{r transform-outliers}
df <- df %>%
  mutate(user_ratings_log = log(user_ratings_total + 1))
```

## 4.3 Categorical Encoding
We group restaurant types and convert them into factors, then generate codes.

```{r encoding}
df <- df %>%
  mutate(types_grouped = case_when(
    types %in% c("Restaurant", "Family restaurant", "Asian restaurant") ~ "Standard_Family",
    types %in% c("Fast food restaurant", "Pizza restaurant", "Hamburger restaurant") ~ "Fast_Food",
    types == "Bangladeshi restaurant" ~ "Traditional_BD",
    types %in% c("Chinese restaurant", "Korean restaurant", "Japanese restaurant", "Thai restaurant") ~ "East_Asian",
    types %in% c("Sweets and dessert buffet", "Bakery", "Cafe", "Coffee shop", "Ice cream shop") ~ "Cafe_Dessert",
    types %in% c("Biryani restaurant", "Kebab shop") ~ "Biryani_Kebab",
    TRUE ~ "Other_Specialty"
  ))

# Convert to Factors
df$division <- as.factor(df$division)
df$types_grouped <- as.factor(df$types_grouped)

# Generate numeric codes
df$division_code <- as.numeric(df$division)
df$type_code <- as.numeric(df$types_grouped)

# Manual One-Hot Encoding for models
divisions <- unique(df$division)
for (div in divisions) {
  col_name <- paste0("div_", gsub(" ", "_", div))
  df[[col_name]] <- ifelse(df$division == div, 1, 0)
}

types_cats <- unique(df$types_grouped)
for (t_cat in types_cats) {
  col_name <- paste0("type_", t_cat)
  df[[col_name]] <- ifelse(df$types_grouped == t_cat, 1, 0)
}
```

## 4.4 Feature Engineering
We create two new features:
1. `competitor_density`: Number of restaurants in the same division.
2. `service_diversity`: Total count of service options offered.

```{r feature-engineering}
# Feature 1: Competitor Density
df <- df %>%
  group_by(division) %>%
  mutate(competitor_density = n()) %>%
  ungroup()

# Feature 2: Service Diversity
df <- df %>%
  mutate(service_diversity = ifelse(is.na(`Service options`), 0, str_count(`Service options`, ",") + 1))
```

**Justification:**
* **`competitor_density`:** Helps the model see how many other food options are in the same division.
* **`service_diversity`:** A higher count means the restaurant is more flexible (e.g. offers delivery, takeaway, dine-in), which can attract more customers.

## 4.5 Numeric Scaling
We standardize the engineered variables using `scale()`, and manually normalize the coordinates using the Min-Max formula.

```{r scaling}
# Standardization
df$competitor_density_std <- scale(df$competitor_density)
df$service_diversity_std <- scale(df$service_diversity)

# Min-Max Scaling (Normalization)
df$lat_norm <- (df$lat - min(df$lat)) / (max(df$lat) - min(df$lat))
df$lng_norm <- (df$lng - min(df$lng)) / (max(df$lng) - min(df$lng))
```

## 4.6 Feature Selection
We check for variables with near-zero variance and remove highly correlated variables.

```{r feature-selection}
features_to_check <- df %>%
  select(lat_norm, lng_norm, competitor_density_std, service_diversity_std, 
         starts_with("div_"), starts_with("type_"))

# Near-Zero Variance filter
nzv_metrics <- caret::nearZeroVar(features_to_check, saveMetrics = TRUE)
low_variance_vars <- rownames(nzv_metrics[nzv_metrics$nzv == TRUE, ])

cat("Low Variance removed:", ifelse(length(low_variance_vars) == 0, "None", paste(low_variance_vars, collapse = ", ")), "\n")

# Correlation filter
cor_mat <- cor(features_to_check %>% select(-any_of(low_variance_vars)))
high_cor_indices <- caret::findCorrelation(cor_mat, cutoff = 0.85)
high_cor_vars <- colnames(cor_mat)[high_cor_indices]

cat("Highly correlated removed:", ifelse(length(high_cor_vars) == 0, "None", paste(high_cor_vars, collapse = ", ")), "\n")
```

```{r final-dataset}
# Build final dataset
df_clean <- df %>%
  select(is_high_potential, lat_norm, lng_norm, competitor_density_std, service_diversity_std, 
         starts_with("div_"), starts_with("type_")) %>%
  select(-any_of(c(low_variance_vars, high_cor_vars)))

# Save preprocessed dataset
save(df_clean, file = "df_clean.RData")

dim(df_clean)
```

**Decision:** We checked for highly correlated features. The `competitor_density` feature was removed because it was highly correlated, leaving a final clean dataset of **39,409 rows and 14 columns** for modeling in Phase 2.

---

*End of Phase 1 — Midterm Submission*
